Make any stupid plot you want

### 2021-02-24 draft/sections/intro.tex
done for now w/ gog cause this took 2 hours

In other words this model is not intended to be evaluative, it is intended to be a reference specification for visualization library API. To make this model as implementation independent as possible, we propose fairly general mathematical abstractions of the data container such that we do not need to assume the data has any specific structure, such as a relational database. We reuse this structure for the graphic as that allows us to specifically discuss how structure is preserved. We take a functional approach because that gives implementers the flexibility to carry as little or as much state as they need and allows us to specify the constraints on these functions without having to know anything about the implementation or any state that these functions may be carrying. As with the other mathematical formalisms of visualization, we factor out the rendering into a separate stage; but our framework describes how these rendering instructions are generated.

@@

A commonly cited alternative is the family of tools built on top of Wilkenson's Grammar of Graphics (GoG) \cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}. The GoG family of tools are very strongly built on the assumption that data is a relational framework; this is explicit in Mackinlay's A Presentation Tool (APT)\cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987} formalism on which GoG is build. This framework is very popular in the data visualization community because it lends itself to a declarative interface \cite{heerDeclarative2010} which allows end users to describe the visualization they are trying to create. Grammar oriented approaches describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building the elements. A clear example of this distinction is that the GoG grammar includes computation and aggregation of the table as part of the grammar, while we propose that computations are specific to domains and only try to describe them when they are specifically part of the visual encoding - for example mapping data to a color.

assumptions on each specific component of the visualization, which in turn allows for building composable

>>>>

In other words this model is not intended to be evaluative, it is intended to be a reference specification for visualization library API. To make this model as implementation independent as possible, we propose fairly general mathematical abstractions of the data container such that we do not need to assume the data has any specific structure, such as a relational database. We reuse this structure for the graphic as that allows us to specifically discuss how structure is preserved. We take a functional approach because functional paradigms encourage writing APIs that are flexible, concise and predictable due to the lack of side effects \cite{loudenProgrammingLanguagesPrinciples2010}. Furthermore, by structuring the API in terms of composition of the smallest units of transformation for which we can define correctness, a functional paradigm naturally leads to a library of highly modular components that are composable in such a way that by definition the composition is also correct. This allows us to ensure that domain specific visualizations built on top of these components are also correct without needing knowledge of the domain. As with the other mathematical formalisms of visualization, we factor out the rendering into a separate stage; but, our framework describes how these rendering instructions are generated.

@@

A commonly cited alternative is the family of tools built on top of Wilkenson's Grammar of Graphics (GoG) \cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}. The GoG family of libraries are very strongly built on the assumption that data is a relational table, which restricts the native set of visualization types the libraries support. Since the data space and graphic space is very well defined in this grammar, it lends itself to a declarative interface \cite{heerDeclarative2010} that allows end users to describe the visualization they are trying to create. Grammar oriented approaches describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building those elements. An example of this distinction is that the GoG grammar includes computation and aggregation of the table as part of the grammar, while we propose that most computations are specific to domains and only try to describe them when they are specifically part of the visual encoding - for example mapping data to a color. Disentangling the computation from the visual transforms allows us to determine whether the visualization library needs to handle them or if they can be more efficiently computed by the data container.
